---
params:
  config: "config/config.yaml"
  root: ""
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    toc-location: left
    embed-resources: true
    fig-align: center
    tbl-cap-location: top
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
# ---------------------------------------------------------------------------
# workflow/report/REPORT.qmd — the cohort-agnostic run report.
#
# This document contains NO cohort-specific content. Everything it says comes
# from (a) the active config and (b) which output files the run actually
# produced. Cluster names, group names, pair names and role columns are read
# at render time, never written in. A stage that did not run gets a short note
# saying why — the same graceful-degradation discipline the rules follow.
#
# Rendered by `snakemake --configfile <cfg> report`; see
# workflow/rules/99_report.smk, which passes the config path and repo root in
# as Quarto parameters.
# ---------------------------------------------------------------------------
suppressPackageStartupMessages({
  library(yaml)
  library(knitr)
})

`%||%` <- function(a, b) {
  if (is.null(a) || length(a) == 0) return(b)
  if (length(a) == 1 && is.na(a)) return(b)
  a
}

CFG_PATH <- normalizePath(params$config, mustWork = TRUE)
CFG      <- yaml::read_yaml(CFG_PATH)

# Config paths are relative to the workspace root — the directory the pipeline
# runs from — not to this file. The rule passes it explicitly; the fallback
# only covers running this document by hand.
ROOT <- if (nzchar(params$root %||% "")) {
  normalizePath(params$root, mustWork = TRUE)
} else {
  normalizePath(dirname(dirname(CFG_PATH)), mustWork = TRUE)
}

abspath <- function(p) {
  if (is.null(p) || length(p) == 0 || is.na(p) || !nzchar(p)) return(NA_character_)
  if (startsWith(p, "/")) p else file.path(ROOT, p)
}
relpath <- function(p) sub(paste0("^", ROOT, "/"), "", p)

OUT         <- abspath(CFG$paths$outputs)
REP         <- abspath(CFG$paths$reports)
COHORT_NAME <- CFG$cohort$name %||% "cohort"
INPUT_TYPE  <- CFG$cohort$input_type %||% "unknown"

# Roles: the analytical column map. Which roles are set decides which
# role-driven sections can exist at all.
ROLES    <- CFG$metadata$roles %||% list()
role_col <- function(r) {
  v <- ROLES[[r]]
  if (is.null(v) || length(v) == 0 || is.na(v)) NA_character_ else as.character(v)
}
has_role <- function(r) !is.na(role_col(r))

# The {sampleset} axis. Both arms are built whenever clonality.declonalize is
# on, and wherever a result exists for both this report shows them side by
# side rather than picking one — the comparison is the point.
DECLONALIZE <- isTRUE(CFG$clonality$declonalize %||% TRUE)
ARMS        <- c("full", if (DECLONALIZE) "unique")
ARM_LABEL   <- c(full   = "full — every sample",
                 unique = "unique — one per clonal group")

# --- existence helpers -----------------------------------------------------
have <- function(p) {
  length(p) == 1 && !is.na(p) && file.exists(p) && file.size(p) > 0
}
out <- function(...) file.path(OUT, ...)
fig <- function(...) file.path(REP, "figures", ...)

# --- markdown emitters (called from results='asis' chunks) -----------------
md <- function(...) cat(paste0(...), "\n\n", sep = "")

note <- function(...) {
  md("::: {.callout-note appearance=\"simple\"}")
  md("**Not in this run.** ", ...)
  md(":::")
}

# Images go in as base64 data URIs rather than as paths. `embed-resources`
# alone is not enough: pandoc resolves an absolute image path relative to its
# resource path and ends up looking for "./Users/...", so every figure silently
# fails to embed and the report ships with broken images. A data URI has no
# path to resolve, so it works wherever the document is rendered from — and the
# result is genuinely one self-contained file.
img_uri <- function(path) {
  tryCatch(knitr::image_uri(path), error = function(e) NA_character_)
}

emit_fig <- function(path, caption = "", width = "100%") {
  if (!have(path)) return(invisible(FALSE))
  uri <- img_uri(path)
  if (is.na(uri)) return(invisible(FALSE))
  md("![", caption, "](", uri, "){width=", width, "}")
  invisible(TRUE)
}

# Side-by-side figures for the two sample-set arms. `builder` maps an arm name
# to a path; whichever arms exist are rendered, and a lone arm goes full width.
emit_fig_arms <- function(builder, caption) {
  paths   <- vapply(ARMS, builder, character(1))
  present <- ARMS[vapply(paths, have, logical(1))]
  if (length(present) == 0) return(invisible(FALSE))
  if (length(present) == 1) {
    u <- img_uri(paths[[present]])
    if (is.na(u)) return(invisible(FALSE))
    md("![", caption, " (", ARM_LABEL[[present]], ")](", u, "){width=100%}")
  } else {
    md("::: {layout-ncol=", length(present), "}")
    for (a in present) {
      u <- img_uri(paths[[a]])
      if (!is.na(u)) md("![", ARM_LABEL[[a]], "](", u, "){width=100%}")
    }
    md(":::")
    md("*", caption, ".*")
  }
  invisible(TRUE)
}

read_tsv_safe <- function(path) {
  if (!have(path)) return(NULL)
  tryCatch(
    utils::read.delim(path, sep = "\t", header = TRUE, check.names = FALSE,
                      stringsAsFactors = FALSE, comment.char = "", quote = ""),
    error = function(e) NULL
  )
}

emit_tbl <- function(path, caption = NULL, n = 15) {
  d <- read_tsv_safe(path)
  if (is.null(d) || nrow(d) == 0) return(invisible(FALSE))
  total <- nrow(d)
  print(knitr::kable(utils::head(d, n), caption = caption,
                     format = "pipe", row.names = FALSE))
  cat("\n\n")
  if (total > n)
    md("*Showing ", n, " of ", format(total, big.mark = ","),
       " rows. Full table: `", relpath(path), "`.*")
  invisible(TRUE)
}

emit_df <- function(d, caption = NULL) {
  if (is.null(d) || nrow(d) == 0) return(invisible(FALSE))
  print(knitr::kable(d, caption = caption, format = "pipe", row.names = FALSE))
  cat("\n\n")
  invisible(TRUE)
}

count_lines <- function(path, header = TRUE) {
  if (!have(path)) return(NA_integer_)
  n <- length(readLines(path, warn = FALSE))
  if (header) max(n - 1L, 0L) else n
}
fmt <- function(x) if (is.na(x)) "—" else format(x, big.mark = ",")

# Cluster, pair and model names are DERIVED, never hardcoded — they come from
# whatever this run's labelling produced.
clusters_for <- function(arm) {
  d <- read_tsv_safe(out("structure", arm, "admix_clusters.tsv"))
  if (is.null(d) || !("Cluster" %in% names(d))) return(character(0))
  sort(unique(as.character(d$Cluster)))
}
CLUSTERS <- clusters_for("full")

pairs_for <- function(arm) {
  d <- out("introgression", arm, "pairs")
  if (!dir.exists(d)) return(character(0))
  sort(sub("\\.tsv$", "", basename(list.files(d, pattern = "\\.tsv$"))))
}

FOCAL            <- CFG$introgression$focal_group %||% NA_character_
SELECTION_MODELS <- CFG$selection$models %||% list()
```

```{r title}
#| results: asis
md("# ", COHORT_NAME, " — population genetics report")
md("Generated ", format(Sys.time(), "%Y-%m-%d %H:%M"), " by **ZOOMAL-Flow** from `",
   relpath(CFG_PATH), "`.")
md("Every section is assembled from what this run actually produced. Stages that ",
   "did not run are marked rather than omitted, and group, cluster and column names ",
   "are read from your config and your outputs — nothing here is written for a ",
   "particular cohort.")
```

## Overview

```{r overview}
#| results: asis
n_var <- if (have(out("qc", "variant_count.txt")))
  trimws(readLines(out("qc", "variant_count.txt"), warn = FALSE)[1]) else NA_character_

emit_df(data.frame(
  Quantity = c("Cohort", "Input type", "Samples in the VCF", "Samples with metadata",
               "Samples after control filtering", "Variants after QC",
               "Samples into structure analysis", "Sample-set arms built"),
  Value = c(COHORT_NAME, INPUT_TYPE,
            fmt(count_lines(out("setup", "vcf_samples.txt"), header = FALSE)),
            fmt(count_lines(out("metadata", "samples.tsv"))),
            fmt(count_lines(out("qc", "keep_samples.txt"), header = FALSE)),
            if (is.na(n_var)) "—" else n_var,
            fmt(count_lines(out("structure", "full", "cleaned.fam"), header = FALSE)),
            paste(ARMS, collapse = " + ")),
  stringsAsFactors = FALSE, check.names = FALSE))
```

### Metadata roles

The pipeline works off *roles*, not column names: you map your table's columns onto
the analytical roles below, and any role you leave unset disables the analyses that
depend on it instead of failing.

```{r roles}
#| results: asis
role_effect <- c(
  sample_id    = "required — identifies every sample",
  group        = "group-faceted plots, group-based selection filters",
  geography    = "geography facets, the map, geography summaries",
  country      = "country facets and summaries",
  host         = "recorded only; no analysis depends on it",
  date         = "temporal analyses (the clonal-group timeline)",
  case_control = "case/control contrasts in the selection scan"
)
emit_df(data.frame(
  Role          = names(role_effect),
  `Your column` = vapply(names(role_effect), function(r) {
                    v <- role_col(r); if (is.na(v)) "—" else paste0("`", v, "`") },
                    character(1)),
  Status        = vapply(names(role_effect),
                         function(r) if (has_role(r)) "mapped" else "not set", character(1)),
  Enables       = unname(role_effect),
  check.names = FALSE, stringsAsFactors = FALSE))

unset <- names(role_effect)[!vapply(names(role_effect), has_role, logical(1))]
if (length(unset))
  md("Unset in this run: ", paste0("`", unset, "`", collapse = ", "),
     ". The analyses those drive are skipped below.")
```

### Clonality handling

```{r clonality-note}
#| results: asis
if (DECLONALIZE) {
  md("Clonal samples are near-identical genotypes. Counting each as independent ",
     "evidence is pseudo-replication, and it biases anything that is fundamentally a ",
     "frequency or a count — allele frequencies, ancestry proportions, per-group ",
     "polyclonality rates, introgression window support. So every frequency stage is ",
     "built **twice**:")
  md("- **full** — every sample. The operational backbone: clonal detection and the ",
     "cluster definitions key off this arm.")
  md("- **unique** — one representative per clonal group, plus every sample that is in ",
     "no clonal group. The comparison arm.")
  md("Wherever a result exists for both, this report shows them side by side. A claim ",
     "that looks the same in both arms survives de-clonalization; one that changes did ",
     "not. See [Full vs de-clonalized](#full-vs-de-clonalized) for the numbers.")
} else {
  note("`clonality.declonalize` is **false**, so only the `full` sample set was built ",
       "and there is no de-clonalized comparison. Frequency-based results below ",
       "therefore count clonal replicates as independent samples.")
}
```

## Quality control

```{r qc}
#| results: asis
qc_figs <- list(
  c("maf_density_overall.png",     "Minor-allele-frequency distribution across retained variants"),
  c("maf_density_by_chrom.png",    "Minor-allele frequency per contig"),
  c("miss_density_per_sample.png", "Missing-genotype rate per sample"),
  c("miss_density_per_variant.png","Missing-genotype rate per variant")
)
any_qc <- FALSE
for (f in qc_figs) if (emit_fig(fig(f[1]), f[2])) any_qc <- TRUE

if (!any_qc) {
  note("No QC figures were produced. These are generated on the WGS path; ",
       "`cohort.input_type` here is `", INPUT_TYPE, "`.")
} else {
  md("These are the sanity check before anything is interpreted. A missingness ",
     "distribution with a long right tail means samples or sites are being carried ",
     "that the structure-stage thresholds (`structure.max_sample_missing` = ",
     CFG$structure$max_sample_missing %||% "—", ", `structure.max_variant_missing` = ",
     CFG$structure$max_variant_missing %||% "—", ") will later drop.")
}
```

## Multiplicity of infection

```{r moi-intro}
#| results: asis
if (have(out("moi", "fws_MOI.tsv"))) {
  md("**F~ws~** measures how much genetic diversity a single infection contains ",
     "relative to the population it came from. It runs from 0 to 1: near 1 means the ",
     "infection is effectively one clone, and lower values mean several distinct ",
     "parasite genotypes are present in the one host — a *polyclonal* infection. It is ",
     "estimated from within-sample allele balance, so it needs allele-depth data, not ",
     "just genotype calls.")
  md("It matters here for a concrete reason: ADMIXTURE, PCA and IBD all assume one ",
     "genotype per sample, so samples below `moi.fws_exclusion_cutoff` (",
     CFG$moi$fws_exclusion_cutoff %||% "—", ") are excluded before those stages.")
  md("F~ws~ was scored for **", fmt(count_lines(out("moi", "fws_MOI.tsv"))),
     "** samples; **", fmt(count_lines(out("moi", "exclude_high_moi.txt"), header = FALSE)),
     "** were excluded as too polyclonal. (`moi.fws_polyclonal_cutoff` = ",
     CFG$moi$fws_polyclonal_cutoff %||% "—",
     " is the separate, higher threshold used only for *reporting* polyclonality rates.)")
} else {
  note("The MOI stage produced no `fws_MOI.tsv`.")
}
```

```{r moi-figs}
#| results: asis
for (r in c("country", "geography")) {
  if (!has_role(r)) next
  emit_fig(fig(paste0("fws_density_by_", r, ".png")),
           paste0("F~ws~ distribution by ", r, " (role column `", role_col(r), "`)"))
}
```

```{r moi-tables}
#| results: asis
if (!has_role("country") && !has_role("geography")) {
  note("Neither the `country` nor the `geography` role is mapped, so there are no ",
       "grouped polyclonality summaries. F~ws~ is still scored per sample.")
} else {
  for (r in c("country", "geography")) {
    if (!has_role(r)) next
    for (a in ARMS) {
      emit_tbl(out("moi", a, paste0("fws_by_", r, ".tsv")),
               caption = paste0("Polyclonality by ", r, " — ", ARM_LABEL[[a]]), n = 20)
    }
  }
  if (DECLONALIZE)
    md("The two arms are shown one after the other. A per-group polyclonality rate ",
       "that moves between them was being driven by clonal replicates within that group.")
}
```

## Population structure

```{r structure-intro}
#| results: asis
if (!length(CLUSTERS)) {
  note("No `admix_clusters.tsv` was found, so the structure stage did not complete.")
} else {
  md("Three complementary views of one question — how do these samples group ",
     "genetically? **ADMIXTURE** models each sample as a mixture of *K* ancestral ",
     "populations and reports the proportions. **PCA** projects the genotypes onto the ",
     "axes of greatest variation, with no population model behind it. A ",
     "**neighbour-joining tree** summarises the pairwise genetic-distance matrix. They ",
     "should broadly agree; where they disagree, understand why before going further.")
  md("Clusters recovered in this run: ", paste0("`", CLUSTERS, "`", collapse = ", "),
     " (labelling mode `", CFG$structure$cluster_labelling %||% "—", "`).")

  # Was LD-pruning actually applied? On a small cohort it is skipped, and that
  # materially changes how much weight the structure results carry — so say so
  # here rather than leaving it in a log nobody opens.
  for (a in ARMS) {
    st <- out("structure", a, "ld_prune_status.txt")
    if (!have(st)) next
    line <- readLines(st, warn = FALSE)[1]
    if (!grepl("^SKIPPED", line)) next
    n_s <- sub(".*n_samples=([0-9]+).*", "\\1", line)
    thr <- sub(".*threshold=([0-9]+).*", "\\1", line)
    md("::: {.callout-warning}")
    md("## LD-pruning was skipped for the `", a, "` arm")
    md("This sample set has **", n_s, " samples**, below ",
       "`structure.min_samples_for_ld_prune` (", thr, "). PLINK2 will not ",
       "estimate linkage disequilibrium from fewer than 50 samples, and an ",
       "estimate from that few haplotypes would be mostly noise, so the ",
       "ADMIXTURE / PCA / tree results below were computed on the **unpruned** ",
       "variant set.")
    md("Linked variants are correlated evidence, so they can inflate apparent ",
       "structure: a cluster may be driven by one long haplotype rather than by ",
       "genome-wide ancestry. Treat the structure results — and the K selection ",
       "in particular — as indicative rather than settled.")
    md(":::")
  }
}
```

```{r structure-k}
#| results: asis
if (length(CLUSTERS)) {
  krows <- do.call(rbind, lapply(ARMS, function(a) {
    bk <- out("structure", a, "best_k.txt")
    if (!have(bk)) return(NULL)
    cv <- read_tsv_safe(out("structure", a, "admixture", "cv_error.tsv"))
    data.frame(
      `Sample set`      = unname(ARM_LABEL[[a]]),
      `Selected K`      = trimws(readLines(bk, warn = FALSE)[1]),
      `K scanned`       = if (!is.null(cv)) paste0(min(cv[[1]]), "–", max(cv[[1]])) else "—",
      `Lowest CV error` = if (!is.null(cv)) sprintf("%.5f", min(cv[[2]])) else "—",
      check.names = FALSE, stringsAsFactors = FALSE)
  }))
  emit_df(krows, "Model selection: K chosen by cross-validation error")
  md("Cross-validation error is a model-selection *diagnostic*, not a test. The K with ",
     "the lowest CV error is the most defensible choice, but the curve is often flat ",
     "near its minimum, and ADMIXTURE's threaded cross-validation is not ",
     "bit-reproducible run to run. Treat small differences in CV error as noise rather ",
     "than as evidence for one K over its neighbour.")
  emit_fig_arms(function(a) fig(a, "admixture_cv.png"), "Cross-validation error against K")
}
```

```{r structure-figs}
#| results: asis
if (length(CLUSTERS)) {
  emit_fig_arms(function(a) fig(a, "admixture_bars.png"),
                "ADMIXTURE ancestry proportions per sample at the selected K")
  for (r in c("country", "geography")) {
    if (!has_role(r)) next
    local({
      rr <- r
      emit_fig_arms(function(a) fig(a, paste0("admixture_bars_by_", rr, ".png")),
                    paste0("Ancestry proportions faceted by ", rr,
                           " (role column `", role_col(rr), "`)"))
    })
  }
  emit_fig_arms(function(a) fig(a, "pca.png"), "Principal components, coloured by cluster")
  if (has_role("geography"))
    emit_fig_arms(function(a) fig(a, "pca_by_geography.png"),
                  paste0("Principal components, coloured by `", role_col("geography"), "`"))
  emit_fig_arms(function(a) fig(a, "njt.png"),
                "Neighbour-joining tree of pairwise genetic distance")
}
```

```{r structure-sizes}
#| results: asis
if (length(CLUSTERS)) {
  sizes <- do.call(rbind, lapply(ARMS, function(a) {
    d <- read_tsv_safe(out("structure", a, "admix_clusters.tsv"))
    if (is.null(d) || !("Cluster" %in% names(d))) return(NULL)
    tb <- as.data.frame(table(d$Cluster), stringsAsFactors = FALSE)
    names(tb) <- c("Cluster", "n")
    data.frame(`Sample set` = unname(ARM_LABEL[[a]]), tb,
               check.names = FALSE, stringsAsFactors = FALSE)
  }))
  emit_df(sizes, "Cluster membership, by sample set")
}
```

```{r structure-map}
#| results: asis
if (!emit_fig(fig("province_map.png"),
              paste0("Sample origins by `", role_col("geography") %||% "geography", "`"))) {
  if (is.null(CFG$metadata$gis))
    note("No map: `metadata.gis` is not set, so there are no per-sample coordinates to ",
         "place. Point it at a table with sample-ID, latitude and longitude columns to ",
         "enable it.")
}
```

## Identity by descent and clonality

```{r ibd-intro}
#| results: asis
cc <- out("ibd", "clonal_clusters.tsv")
if (have(cc)) {
  md("Two samples are **identical by descent** (IBD) over a genomic segment when they ",
     "inherited it from a shared ancestor. hmmIBD walks each pair along the genome with ",
     "a hidden Markov model and reports the fraction of sites in an IBD state. Pairs ",
     "above `ibd.clonal_ibd_threshold` (", CFG$ibd$clonal_ibd_threshold %||% "—",
     ") are treated as the same clonal genotype, and connected components of those pairs ",
     "form clonal groups.")
  md("IBD is computed **within** each cluster — the model assumes a single population's ",
     "allele frequencies — and only for clusters with at least `ibd.min_cluster_n` (",
     CFG$ibd$min_cluster_n %||% "—", ") members.")
} else {
  note("No IBD output was found: the stage did not run, or no cluster reached ",
       "`ibd.min_cluster_n` (", CFG$ibd$min_cluster_n %||% "—", ") members.")
}
```

```{r ibd-tables}
#| results: asis
audit <- out("clonality", "declonalization_audit.tsv")
ad <- read_tsv_safe(audit)
if (!is.null(ad) && nrow(ad) > 0) {
  emit_tbl(audit, "Clonal groups, with the representative retained in the `unique` arm", n = 25)
  if ("straddles_clusters" %in% names(ad)) {
    n_str <- sum(tolower(as.character(ad$straddles_clusters)) %in% c("true", "yes", "1"),
                 na.rm = TRUE)
    if (n_str > 0)
      md("**", n_str, " clonal group(s) straddle more than one cluster.** Worth a look: ",
         "near-identical genotypes landing in different clusters usually means either a ",
         "cluster boundary in the wrong place or duplicated material in the cohort.")
  }
} else if (have(audit)) {
  md("**No clonal groups were detected** at the configured threshold. The `unique` arm ",
     "is therefore identical to `full`, and every difference between the two arms below ",
     "should be nil — which is itself worth confirming rather than assuming.")
} else if (DECLONALIZE) {
  md("No clonality audit table was written.")
}
```

```{r ibd-figs}
#| results: asis
conn <- if (dir.exists(fig()))
  list.files(fig(), pattern = "^ibd_connectivity_.*\\.png$", full.names = TRUE) else character(0)
if (length(conn)) {
  md("Each panel is one cluster's IBD network: nodes are samples, edges join pairs ",
     "sharing more IBD than the plotting threshold. Dense cliques are clonal expansions; ",
     "isolated nodes are unrelated infections.")
  for (p in sort(conn))
    emit_fig(p, paste0("IBD connectivity — cluster `",
                       sub("^ibd_connectivity_(.*)\\.png$", "\\1", basename(p)), "`"))
}

if (has_role("date")) {
  emit_fig(fig("clonal_temporal.png"),
           paste0("Clonal groups over time (role column `", role_col("date"), "`)"))
} else {
  note("No clonal timeline: the `date` role is not mapped, so there are no collection ",
       "dates to plot clonal groups against.")
}
```

## Introgression

```{r intro-intro}
#| results: asis
have_intro <- any(vapply(ARMS,
  function(a) have(out("introgression", a, "windows_by_cluster.tsv")), logical(1)))

if (!have_intro) {
  note("The introgression stage produced no window tables: it did not run, or no cluster ",
       "reached `introgression.min_cluster_n` (",
       CFG$introgression$min_cluster_n %||% "—", ").")
} else {
  md("This stage asks, per genomic window: does a sample carry the genetic signature of ",
     "a cluster *other* than the one it is assigned to? It works per cluster pair, ",
     "measuring each sample-window's distance to both clusters' consensus alleles and ",
     "flagging windows where a sample sits in the other cluster's territory rather than ",
     "its own.")
  ps <- pairs_for("full")
  md("Window size ", fmt(CFG$introgression$window_size_bp %||% NA), " bp; at least ",
     CFG$introgression$min_snps_per_window %||% "—", " SNPs per window; detection rule `",
     CFG$introgression$detection_rule %||% "—", "`.",
     if (length(ps)) paste0(" Cluster pairs compared: ",
                            paste0("`", ps, "`", collapse = ", "), ".") else "")
  md("::: {.callout-important}")
  md("## Cluster-level and focal-level are separate questions")
  md("They are two tests at two scales, each with its own null. The cluster-level support ",
     "floor should be **derived by permutation on this cohort** against an explicit FDR ",
     "target — a number inherited from another study is not valid here. The focal test ",
     "has its own size-preserving permutation null, scaled to the subgroup. A subgroup ",
     "can never clear a floor calibrated on a cluster many times its size, so filtering ",
     "the focal question through the cluster floor is a mis-posed test, not a threshold ",
     "to tune. See `docs/introgression_analysis_spec.md`.")
  md(":::")
}
```

```{r intro-tables}
#| results: asis
if (have_intro) {
  for (a in ARMS) {
    p <- out("introgression", a, "windows_by_cluster.tsv")
    if (!have(p)) next
    md("**", ARM_LABEL[[a]], "** — ", fmt(count_lines(p)), " windows retained.")
    emit_tbl(p, caption = paste0("Introgressed windows by cluster — ", ARM_LABEL[[a]]), n = 12)
  }
  emit_fig_arms(function(a) fig(a, "introgression_shoulder.png"),
                "Per-sample introgression across the genome")
}
```

```{r intro-focal}
#| results: asis
if (!is.na(FOCAL)) {
  md("### Focal-group enrichment — `", FOCAL, "`")
  md("This asks whether introgression in a window is **enriched** in the focal subgroup ",
     "relative to the rest of its own cluster, under a size-preserving label permutation ",
     "with Benjamini–Hochberg control at FDR ", CFG$introgression$focal_fdr %||% "—",
     ". This is the testable focal claim — cite this table.")
  for (a in ARMS) {
    p <- out("introgression", a, paste0("focal_", FOCAL, "_enriched_windows.tsv"))
    if (!have(p)) next
    if (count_lines(p) == 0)
      md("**", ARM_LABEL[[a]], "** — no windows reached significance.")
    else
      emit_tbl(p, caption = paste0("Enriched windows in `", FOCAL, "` — ", ARM_LABEL[[a]]), n = 15)
  }
  emit_fig_arms(function(a) fig(a, paste0("introgression_focal_", FOCAL, ".png")),
                paste0("Focal enrichment — `", FOCAL, "`"))
  if (DECLONALIZE)
    md("Compare the arms directly. A focal result present in `full` and absent in ",
       "`unique` was carried by clonal replicates rather than by independent infections, ",
       "and does not survive de-clonalization.")
  md("::: {.callout-warning}")
  md("## The `unique_windows_in_", FOCAL, "_*.tsv` table is descriptive only")
  md("It is a raw set difference: no null, no p-value, no multiple-testing control, and ",
     "a single background sample carrying a window removes it from the set. It is ",
     "superseded by the enrichment table above and must not be cited as a result.")
  md(":::")
} else if (have_intro) {
  note("`introgression.focal_group` is not set, so no focal-enrichment test ran. Set it ",
       "to a value of the `", CFG$introgression$focal_role %||% "geography",
       "` role to ask whether introgression is enriched in a particular subgroup.")
}
```

## Selection

```{r selection}
#| results: asis
if (length(SELECTION_MODELS) == 0) {
  note("`selection.models` is empty, so no selection scan ran. An iHS scan needs a ",
       "case/control contrast, which is a cohort-specific hypothesis — add a model with ",
       "`case_filter` and `control_filter` expressions to enable it.")
} else {
  md("The **integrated haplotype score** (iHS) looks for alleles that rose in frequency ",
     "fast enough to carry an unusually long haplotype with them — the signature of ",
     "recent positive selection. One scan is run per contrast.")
  for (m in SELECTION_MODELS) {
    nm <- m$name
    md("### ", nm)
    md("Cases: `", m$case_filter, "` · Controls: `", m$control_filter, "`")
    got <- FALSE
    for (a in ARMS)
      if (emit_tbl(out("selection", a, nm, "ihs_extreme_windows.tsv"),
                   paste0("Candidate windows — ", ARM_LABEL[[a]]), n = 15)) got <- TRUE
    local({
      n <- nm
      if (emit_fig_arms(function(a) fig(a, paste0("ihs_", n, ".png")),
                        paste0("iHS scan — ", n))) got <<- TRUE
    })
    if (!got) md("No output was produced for this model.")
  }
}
```

## Full vs de-clonalized {#full-vs-de-clonalized}

```{r comparison}
#| results: asis
if (!DECLONALIZE) {
  note("Only the `full` arm was built (`clonality.declonalize: false`), so there is no ",
       "comparison to show.")
} else {
  md("Each row is one quantity computed both ways. Large changes mark claims that were ",
     "resting on clonal pseudo-replication; near-zero changes mark claims robust to it. ",
     "This table is the reason both arms are built.")
  any_cmp <- FALSE
  for (stage in c("moi", "structure", "introgression", "selection"))
    if (emit_tbl(out(stage, "declonalization_comparison.tsv"),
                 caption = paste0("Full vs de-clonalized — ", stage, " stage"), n = 25))
      any_cmp <- TRUE
  if (!any_cmp) md("No comparison tables were written.")
  emit_fig(fig("declonalization_comparison.png"), "Full vs de-clonalized, across stages")
}
```

## Provenance

```{r provenance}
#| results: asis
md("### Configuration")
md("The full active config, so this report is self-describing:")
cat("``` yaml\n", paste(readLines(CFG_PATH, warn = FALSE), collapse = "\n"), "\n```\n\n", sep = "")

md("### Software")
ver1 <- function(cmd, args = "--version") {
  o <- suppressWarnings(tryCatch(
    system2(cmd, args, stdout = TRUE, stderr = FALSE),
    error = function(e) character(0), warning = function(w) character(0)))
  if (length(o) == 0) "not found" else trimws(o[1])
}
emit_df(data.frame(
  Tool    = c("R", "bcftools", "plink", "plink2", "snakemake", "quarto"),
  Version = c(paste(R.version$major, R.version$minor, sep = "."),
              ver1("bcftools"), ver1("plink"), ver1("plink2"),
              ver1("snakemake"), ver1("quarto")),
  stringsAsFactors = FALSE))

md("R packages: ", paste(vapply(
  c("tidyverse", "data.table", "SeqArray", "moimix", "rehh", "igraph", "ape"),
  function(p) tryCatch(paste0("`", p, "` ", as.character(packageVersion(p))),
                       error = function(e) paste0("`", p, "` (absent)")),
  character(1)), collapse = ", "), ".")
md("Environment: pinned by `pixi.toml` + `pixi.lock` — see `env/README.md`.")
```
